---
title: "KEGG Pathway Enrichment — Alzheimer's Disease Module Runs"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    theme: flatly
date: "2026-06-25"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE,
                      fig.width = 14, fig.height = 10, dpi = 120)

suppressPackageStartupMessages({
  library(dplyr)
  library(tidyr)
  library(purrr)
  library(readr)
  library(stringr)
  library(tibble)
  library(ggplot2)
  library(scales)
  library(forcats)
})

theme_set(
  theme_bw(base_size = 11) +
    theme(
      panel.grid.minor  = element_blank(),
      strip.background  = element_rect(fill = "#f0f0f0", colour = "grey70"),
      strip.text        = element_text(face = "bold", size = 10),
      legend.position   = "right"
    )
)

# ── Curated list of AD-relevant KEGG pathway codes ───────────────────────────
AD_PATHWAYS <- c(
  "KEGG:05010",  # Alzheimer disease
  "KEGG:05022",  # Pathways of neurodegeneration - multiple diseases
  "KEGG:04724",  # Glutamatergic synapse
  "KEGG:04725",  # Cholinergic synapse
  "KEGG:04721",  # Synaptic vesicle cycle
  "KEGG:04728",  # Dopaminergic synapse
  "KEGG:04726",  # Serotonergic synapse
  "KEGG:04020",  # Calcium signaling pathway
  "KEGG:04010",  # MAPK signaling pathway
  "KEGG:04150",  # mTOR signaling pathway
  "KEGG:04210",  # Apoptosis
  "KEGG:04140",  # Autophagy - animal
  "KEGG:04141",  # Protein processing in endoplasmic reticulum
  "KEGG:04144",  # Endocytosis
  "KEGG:04310",  # Wnt signaling pathway
  "KEGG:04330",  # Notch signaling pathway
  "KEGG:04110",  # Cell cycle
  "KEGG:05016",  # Huntington disease
  "KEGG:05014",  # Amyotrophic lateral sclerosis
  "KEGG:05012"   # Parkinson disease
)

# ── Neurodegenerative disease pathways (KEGG 05xxx disease section) ───────────
NEURODEGEN_PATHWAYS <- c(
  "KEGG:05010",  # Alzheimer disease
  "KEGG:05022",  # Pathways of neurodegeneration - multiple diseases
  "KEGG:05012",  # Parkinson disease
  "KEGG:05016",  # Huntington disease
  "KEGG:05014",  # Amyotrophic lateral sclerosis
  "KEGG:05017",  # Spinocerebellar ataxia
  "KEGG:05020"   # Prion disease
)

# ── Colour palettes ───────────────────────────────────────────────────────────
METHOD_COLOURS <- c(
  diamond         = "#d95f02",
  domino          = "#7570b3",
  firstneighbor   = "#1b9e77",
  no_tool         = "#a6761d",
  robust          = "#e7298a",
  robust_bias_aware = "#66a61e",
  rwr             = "#e6ab02"
)

PPI_COLOURS <- c(iid = "#4c78a8", string = "#f58518")

FILTER_COLOURS <- c(filtered = "#2ca25f", unfiltered = "#de77ae")
FILTER_SHAPES  <- c(filtered = 16, unfiltered = 1)

HEATMAP_COLOURS <- colorRampPalette(c("#f7fbff", "#6baed6", "#08519c"))(101)
```

```{r load-data}
# ── Metadata parser ───────────────────────────────────────────────────────────
parse_module <- function(module_id) {
  method <- str_extract(
    module_id,
    "(robust_bias_aware|diamond|domino|firstneighbor|no_tool|robust|rwr)$"
  )
  ppi <- case_when(
    str_detect(module_id, "\\.iid\\.") ~ "iid",
    str_detect(module_id, "\\.string\\.") ~ "string",
    TRUE ~ "other"
  )
  filter_status <- if_else(
    str_detect(module_id, "\\.Brain_Hippocampus\\."),
    "filtered", "unfiltered"
  )
  short_label <- paste0(
    toupper(ppi), " · ",
    str_replace_all(method, "_", "-"), " · ",
    if_else(filter_status == "filtered", "flt", "unfl")
  )
  tibble(module_id, ppi, method, filter_status, short_label)
}

# ── Read one run directory ────────────────────────────────────────────────────
read_run <- function(run_dir, threshold) {
  kegg_files <- list.files(
    file.path(run_dir, "evaluation", "gprofiler"),
    pattern = "\\.KEGG\\.sub_enriched_pathways\\.tsv$",
    recursive = TRUE, full.names = TRUE
  )

  map_dfr(kegg_files, function(path) {
    module_id <- basename(dirname(path))
    read_tsv(path, show_col_types = FALSE) %>%
      mutate(module_id = module_id)
  }) %>%
    filter(Pathway_name != "KEGG root term") %>%
    mutate(threshold = threshold)
}

# ── Load all three thresholds ─────────────────────────────────────────────────
run_meta <- tibble(
  run_dir   = sort(list.dirs("alzheimer_symbol_runs", recursive = FALSE,
                             full.names = TRUE)),
  threshold = str_extract(basename(run_dir), "[0-9.]+$")
) %>%
  filter(!is.na(threshold))

kegg_all <- map2_dfr(run_meta$run_dir, run_meta$threshold, read_run) %>%
  left_join(
    map_dfr(unique(.$module_id), parse_module),
    by = "module_id"
  ) %>%
  mutate(
    neg_log10_padj        = -log10(Padj),
    is_ad_pathway         = Pathway_code %in% AD_PATHWAYS,
    is_neurodegen_pathway = Pathway_code %in% NEURODEGEN_PATHWAYS,
    threshold      = factor(threshold, levels = sort(unique(as.numeric(threshold))))
  )

# ── Module ordering: PPI → method → filter_status ────────────────────────────
module_order_tbl <- kegg_all %>%
  distinct(module_id, ppi, method, filter_status, short_label) %>%
  mutate(
    ppi    = factor(ppi, levels = c("iid", "string")),
    method = factor(method, levels = names(METHOD_COLOURS)),
    filter_status = factor(filter_status, levels = c("filtered", "unfiltered"))
  ) %>%
  arrange(ppi, method, filter_status)
```

## Overview — enriched KEGG pathways per module

How many KEGG pathways (padj < 0.05) does each module return, broken down by PPI network, AMI method, and filtering status?

```{r count-plot, fig.height=7}
count_tbl <- kegg_all %>%
  filter(Padj < 0.05) %>%
  distinct(threshold, module_id, Pathway_code, ppi, method, filter_status) %>%
  count(threshold, module_id, ppi, method, filter_status) %>%
  mutate(
    method = factor(method, levels = names(METHOD_COLOURS)),
    filter_status = factor(filter_status, levels = c("filtered", "unfiltered"))
  )

ggplot(count_tbl,
       aes(x = filter_status, y = n, fill = filter_status)) +
  geom_col(width = 0.65, colour = "white", linewidth = 0.3) +
  geom_text(aes(label = n), vjust = -0.3, size = 3) +
  facet_grid(ppi ~ method + threshold, scales = "free_y") +
  scale_fill_manual(values = FILTER_COLOURS, name = "Filter status") +
  scale_y_continuous(expand = expansion(mult = c(0, 0.15))) +
  labs(
    title  = "Number of significant KEGG pathways (padj < 0.05) per module",
    x      = NULL,
    y      = "# significant pathways"
  ) +
  theme(
    axis.text.x = element_blank(),
    axis.ticks.x = element_blank()
  )
```

---

## AD-relevant pathways — dot plot comparison

Enrichment of the curated set of Alzheimer's-related KEGG pathways across all modules and thresholds. Dot size = fraction of DE genes; colour = –log₁₀(padj); grey = not significant / not detected.

```{r ad-dotplot, fig.height=11, fig.width=16}
ad_data <- kegg_all %>%
  filter(is_ad_pathway) %>%
  mutate(
    sig = Padj < 0.05,
    Fraction_DE = as.numeric(Fraction_DE)
  )

# For each threshold separately so x-axis stays legible
plot_ad_dotplot <- function(thr) {
  d <- ad_data %>% filter(threshold == thr)
  module_levels <- module_order_tbl %>%
    filter(module_id %in% d$module_id) %>%
    pull(short_label)

  pathway_order <- d %>%
    group_by(Pathway_name) %>%
    summarise(best = min(Padj, na.rm = TRUE), .groups = "drop") %>%
    arrange(best) %>%
    pull(Pathway_name)

  d %>%
    mutate(
      short_label   = factor(short_label, levels = module_levels),
      Pathway_name  = factor(Pathway_name, levels = rev(pathway_order)),
      alpha_val     = if_else(sig, 1, 0.25)
    ) %>%
    ggplot(aes(x = short_label, y = Pathway_name)) +
    geom_point(aes(size = Fraction_DE, colour = neg_log10_padj,
                   alpha = alpha_val)) +
    scale_colour_gradientn(
      colours = c("#deebf7", "#9ecae1", "#3182bd", "#08306b"),
      name    = expression(-log[10](p[adj])),
      limits  = c(0, NA)
    ) +
    scale_size_continuous(
      range = c(1.5, 7), name = "Fraction DE",
      labels = percent_format(accuracy = 1)
    ) +
    scale_alpha_identity() +
    facet_grid(. ~ ppi, scales = "free_x", space = "free_x") +
    labs(
      title = paste0("AD-relevant KEGG pathways — threshold ", thr),
      x = NULL, y = NULL
    ) +
    theme(
      axis.text.x = element_text(angle = 50, hjust = 1, size = 8),
      axis.text.y = element_text(size = 9),
      panel.spacing.x = unit(0.4, "cm")
    )
}
```

### Threshold 0.01

```{r ad-dotplot-001, fig.height=10, fig.width=15}
plot_ad_dotplot("0.01")
```

### Threshold 0.1

```{r ad-dotplot-01, fig.height=10, fig.width=15}
plot_ad_dotplot("0.1")
```

### Threshold 1

```{r ad-dotplot-1, fig.height=10, fig.width=15}
plot_ad_dotplot("1")
```

---

## KEGG heatmaps — top pathways per threshold

Top 25 most consistently enriched pathways (lowest mean padj across modules), shown as –log₁₀(padj). AD-relevant pathways are indicated with ★ in the row label. Modules ordered: PPI → AMI method → filter status.

```{r heatmap-fn}
# Build a ggplot2 tile heatmap for one threshold
draw_heatmap <- function(thr, top_n = 25) {
  d <- kegg_all %>%
    filter(threshold == thr, Padj < 0.05)

  if (nrow(d) == 0) {
    message("No significant pathways for threshold ", thr); return(invisible(NULL))
  }

  # Select top_n pathways by lowest mean padj across all modules they appear in
  top_pathways <- d %>%
    group_by(Pathway_name, Pathway_code, is_ad_pathway) %>%
    summarise(
      mean_padj  = mean(Padj, na.rm = TRUE),
      n_modules  = n_distinct(module_id),
      .groups = "drop"
    ) %>%
    arrange(mean_padj) %>%
    slice_head(n = top_n)

  # Row labels: mark AD pathways
  top_pathways <- top_pathways %>%
    mutate(
      row_label = if_else(
        is_ad_pathway,
        paste0("★ ", Pathway_name, " (", Pathway_code, ")"),
        paste0("  ", Pathway_name, " (", Pathway_code, ")")
      )
    )

  # Build the matrix (all modules, even without a hit → NA)
  module_levels <- module_order_tbl %>%
    filter(module_id %in% unique(kegg_all$module_id[kegg_all$threshold == thr])) %>%
    pull(module_id)

  full_grid <- expand_grid(
    module_id    = module_levels,
    Pathway_name = top_pathways$Pathway_name
  )

  plot_data <- full_grid %>%
    left_join(
      d %>% select(module_id, Pathway_name, neg_log10_padj, Padj),
      by = c("module_id", "Pathway_name")
    ) %>%
    left_join(
      module_order_tbl %>% select(module_id, ppi, method, filter_status, short_label),
      by = "module_id"
    ) %>%
    left_join(
      top_pathways %>% select(Pathway_name, row_label, is_ad_pathway, mean_padj),
      by = "Pathway_name"
    ) %>%
    mutate(
      short_label  = factor(short_label,
                            levels = module_order_tbl %>%
                              filter(module_id %in% module_levels) %>%
                              pull(short_label)),
      row_label    = fct_reorder(row_label, -mean_padj),
      filter_status = factor(filter_status, levels = c("filtered", "unfiltered"))
    )

  max_val <- max(plot_data$neg_log10_padj, na.rm = TRUE)

  ggplot(plot_data, aes(x = short_label, y = row_label, fill = neg_log10_padj)) +
    geom_tile(colour = "white", linewidth = 0.35) +
    # Mark AD pathways with a bold left strip
    geom_tile(
      data = plot_data %>% filter(is_ad_pathway),
      aes(x = short_label, y = row_label),
      fill = NA, colour = "#d62728", linewidth = 0.7
    ) +
    scale_fill_gradientn(
      colours  = HEATMAP_COLOURS,
      na.value = "grey92",
      limits   = c(0, max_val),
      name     = expression(-log[10](p[adj]))
    ) +
    facet_grid(. ~ ppi + filter_status, scales = "free_x", space = "free_x") +
    labs(
      title   = paste0("Top ", top_n, " KEGG pathways — threshold ", thr,
                       "  (★ = AD-relevant; red border = AD pathway)"),
      x = NULL, y = NULL
    ) +
    theme(
      axis.text.x  = element_text(angle = 55, hjust = 1, size = 7.5),
      axis.text.y  = element_text(
        size   = 8,
        face   = if_else(
          grepl("^★", levels(plot_data$row_label)),
          "bold", "plain"
        )[as.integer(plot_data$row_label)]
      ),
      panel.spacing.x = unit(0.3, "cm"),
      legend.key.height = unit(1.2, "cm")
    )
}
```

### Threshold 0.01

```{r heatmap-001, fig.height=11, fig.width=16}
draw_heatmap("0.01")
```

### Threshold 0.1

```{r heatmap-01, fig.height=11, fig.width=16}
draw_heatmap("0.1")
```

### Threshold 1

```{r heatmap-1, fig.height=11, fig.width=16}
draw_heatmap("1")
```

---

## Filtered vs. unfiltered — paired pathway enrichment

For each AMI method and PPI network, how does filtering shift the enrichment of AD-relevant pathways? Each line connects a filtered–unfiltered pair for the same method + PPI combination. Points above the diagonal → filtering *increases* enrichment.

```{r paired-scatter, fig.height=8, fig.width=11}
paired <- kegg_all %>%
  filter(is_ad_pathway) %>%
  select(threshold, module_id, Pathway_name, neg_log10_padj,
         ppi, method, filter_status) %>%
  pivot_wider(
    id_cols    = c(threshold, ppi, method, Pathway_name),
    names_from = filter_status,
    values_from = neg_log10_padj,
    values_fill = 0
  ) %>%
  mutate(
    method = factor(method, levels = names(METHOD_COLOURS)),
    diff   = filtered - unfiltered
  )

ggplot(paired, aes(x = unfiltered, y = filtered, colour = method)) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", colour = "grey50") +
  geom_point(alpha = 0.7, size = 2.5, aes(shape = ppi)) +
  scale_colour_manual(values = METHOD_COLOURS, name = "Method") +
  scale_shape_manual(values = c(iid = 16, string = 17), name = "PPI") +
  facet_grid(threshold ~ Pathway_name, labeller = label_wrap_gen(width = 18)) +
  labs(
    title = "Filtered vs. unfiltered enrichment for AD-relevant pathways",
    subtitle = "Points above the dashed line = filtering increased –log₁₀(padj)",
    x = expression("Unfiltered  " * -log[10](p[adj])),
    y = expression("Filtered  " * -log[10](p[adj]))
  ) +
  theme(
    axis.text = element_text(size = 7),
    strip.text = element_text(size = 7.5)
  )
```

---

## Filtering gain — AD pathway enrichment delta

Mean change in –log₁₀(padj) for AD-relevant pathways when applying the Brain Hippocampus filter (filtered minus unfiltered), per AMI method, PPI, and threshold. Positive = filtering helps.

```{r delta-plot, fig.height=7, fig.width=12}
delta_tbl <- kegg_all %>%
  filter(is_ad_pathway) %>%
  select(threshold, ppi, method, filter_status, Pathway_name, neg_log10_padj) %>%
  pivot_wider(
    id_cols    = c(threshold, ppi, method, Pathway_name),
    names_from = filter_status,
    values_from = neg_log10_padj,
    values_fill = 0
  ) %>%
  mutate(delta = filtered - unfiltered) %>%
  group_by(threshold, ppi, method) %>%
  summarise(mean_delta = mean(delta, na.rm = TRUE), .groups = "drop") %>%
  mutate(
    method    = factor(method, levels = names(METHOD_COLOURS)),
    threshold = factor(threshold)
  )

ggplot(delta_tbl,
       aes(x = method, y = mean_delta, fill = method, alpha = threshold)) +
  geom_col(position = position_dodge(width = 0.8), width = 0.7, colour = "white") +
  geom_hline(yintercept = 0, colour = "grey30") +
  scale_fill_manual(values = METHOD_COLOURS, guide = "none") +
  scale_alpha_manual(
    values = c("0.01" = 0.45, "0.1" = 0.7, "1" = 1),
    name   = "Threshold"
  ) +
  facet_wrap(~ ppi, ncol = 2) +
  labs(
    title    = "Mean enrichment gain for AD pathways: filtered − unfiltered",
    subtitle = "Positive = Brain Hippocampus filter increased AD pathway enrichment",
    x = NULL,
    y = expression(Delta ~ mean(-log[10](p[adj])))
  ) +
  theme(axis.text.x = element_text(angle = 35, hjust = 1))
```

---

## Filtering gain — neurodegenerative disease pathways

Same analysis as above but restricted to the seven KEGG neurodegenerative disease pathways (hsa05xxx section): Alzheimer, Parkinson, Huntington, ALS, Spinocerebellar ataxia, Prion disease, and the multi-disease neurodegeneration umbrella pathway. The per-pathway breakdown shows which specific diseases benefit most from the Brain Hippocampus filter.

```{r neurodegen-delta-perpathway, fig.height=8, fig.width=13}
neurodegen_delta <- kegg_all %>%
  filter(is_neurodegen_pathway) %>%
  select(threshold, ppi, method, filter_status, Pathway_name, Pathway_code,
         neg_log10_padj) %>%
  pivot_wider(
    id_cols    = c(threshold, ppi, method, Pathway_name, Pathway_code),
    names_from = filter_status,
    values_from = neg_log10_padj,
    values_fill = 0
  ) %>%
  mutate(
    delta  = filtered - unfiltered,
    method = factor(method, levels = names(METHOD_COLOURS)),
    ppi    = factor(ppi, levels = c("iid", "string")),
    threshold = factor(threshold)
  )

# Short pathway labels
pathway_short <- c(
  "Alzheimer disease"                            = "Alzheimer",
  "Pathways of neurodegeneration - multiple diseases" = "Neurodegeneration\n(multi-disease)",
  "Parkinson disease"                            = "Parkinson",
  "Huntington disease"                           = "Huntington",
  "Amyotrophic lateral sclerosis"                = "ALS",
  "Spinocerebellar ataxia"                       = "Spinocerebellar\nataxia",
  "Prion disease"                                = "Prion disease"
)

neurodegen_delta <- neurodegen_delta %>%
  mutate(pathway_short = recode(Pathway_name, !!!pathway_short))

ggplot(neurodegen_delta,
       aes(x = method, y = delta, fill = method, alpha = threshold)) +
  geom_col(position = position_dodge(width = 0.8), width = 0.7,
           colour = "white", linewidth = 0.2) +
  geom_hline(yintercept = 0, colour = "grey30", linewidth = 0.4) +
  scale_fill_manual(values = METHOD_COLOURS, guide = "none") +
  scale_alpha_manual(
    values = c("0.01" = 0.4, "0.1" = 0.7, "1" = 1),
    name   = "Threshold"
  ) +
  facet_grid(ppi ~ pathway_short) +
  labs(
    title    = "Filtering gain per neurodegenerative disease pathway",
    subtitle = "Positive = Brain Hippocampus filter increased enrichment (filtered − unfiltered)",
    x = NULL,
    y = expression(Delta ~ -log[10](p[adj]))
  ) +
  theme(axis.text.x = element_text(angle = 40, hjust = 1, size = 7.5))
```

### Absolute enrichment — filtered vs. unfiltered side by side

```{r neurodegen-abs, fig.height=7, fig.width=13}
neurodegen_abs <- kegg_all %>%
  filter(is_neurodegen_pathway) %>%
  mutate(
    method        = factor(method, levels = names(METHOD_COLOURS)),
    filter_status = factor(filter_status, levels = c("filtered", "unfiltered")),
    threshold     = factor(threshold),
    pathway_short = recode(Pathway_name, !!!pathway_short)
  )

ggplot(neurodegen_abs,
       aes(x = method, y = neg_log10_padj,
           colour = filter_status, shape = threshold)) +
  geom_point(size = 2.8, alpha = 0.85,
             position = position_dodge(width = 0.6)) +
  geom_hline(yintercept = -log10(0.05), linetype = "dashed",
             colour = "grey50", linewidth = 0.4) +
  scale_colour_manual(values = FILTER_COLOURS, name = "Filter status") +
  scale_shape_manual(values = c("0.01" = 15, "0.1" = 17, "1" = 16),
                     name = "Threshold") +
  facet_grid(ppi ~ pathway_short) +
  labs(
    title    = "Enrichment of neurodegenerative disease pathways across all modules",
    subtitle = "Dashed line = padj 0.05 significance threshold",
    x = NULL,
    y = expression(-log[10](p[adj]))
  ) +
  theme(axis.text.x = element_text(angle = 40, hjust = 1, size = 7.5))
```

---

## Session info

```{r session-info, echo=TRUE}
sessionInfo()
```
